---
title: "Session 5: Point Pattern Analysis II Data Preprocessing"
output: html_notebook
---

For this activity I would like to explore the possibility of using the Scandinavian Bear Project dataset `bear` from the `adehabitatLT` package.

Load the packages needed:
```{r}
library(tidyverse)
library(spatstat)
library(adehabitatLT)
library(lubridate)
```

Load the bear dataset:
```{r}
data(bear)
```

Retrieve the coordinates and dates and place in a dataframe:
```{r}
bear.df <- data.frame(x = bear[[1]]$x, y = bear[[1]]$y, date = bear[[1]]$date)
summary(bear.df)
```

As seen, there are several NA values. Remove incomplete cases:
```{r}
bear.df <- bear.df[complete.cases(bear.df),]
summary(bear.df)
```

Select diurnal and nocturnal observations:
```{r}
bear.df$Time[hour(bear.df$date) <= 7 | hour(bear.df$date) >= 17] <- "Night Time"
bear.df$Time[hour(bear.df$date) >= 7 & hour(bear.df$date) <= 17] <- "Day Time"
```

Convert to factor:
```{r}
bear.df$Time <- as.factor(bear.df$Time)
```

Create a window before converting to `ppp`:
```{r}
W <- owin(c(min(bear.df$x), max(bear.df$x)), c(min(bear.df$y), max(bear.df$y)))
```

Convert to `ppp`:
```{r}
bear.ppp <- as.ppp(X = bear.df, W = W)
summary(bear.ppp)
```

```{r}
plot(split(bear.ppp), which.marks = "Time")
```

Create a new window using `clickpoly()`

Replace window in `ppp` object:
```{r}
junk <- dplyr::select(bear.df, x, y, Time)
junk <- mutate(junk, x = x + runif(nrow(junk), min = 1, max = 2), y = y + runif(nrow(junk), min = 1, max = 2))
bear.ppp <- as.ppp(X = junk, W = Wdw)
summary(bear.ppp)
```

Plot point patterns:
```{r}
plot(split(bear.ppp))
```

Do quadrat analysis:
```{r}
plot(quadratcount(split(bear.ppp), nx = 5, ny = 10))
```

Quadrat test:
```{r}
plot(quadrat.test(split(bear.ppp)$"Night Time", nx = 5, ny = 10))
```


```{r}
kernel_den <- density(split(bear.ppp), sigma = 1000)
plot(kernel_den)
```

Save dataset:
```{r}
save(bear.ppp, file = "Scandinavian Bear Project Example.RData")
```

