---
title: "Data Preprocessing: Practice Statistical Maps II"
output: html_notebook
---

#Introduction

For this activity, I want students to learn more about data types and structures. The plan is that they will use a dataset to create more maps, manipulate information, and see what a list is.

Load libraries:
```{r}
library(tidyverse)
library(ggmap)
library(spdep)
```

First, I imported the dataset used in the social exclusion project of 2008, specifically fast food locations in Toronto.
```{r}
Fast_Food <- read_csv("C:/Antonio/Courses/GEOG 4GA3 - Applied Spatial Analysis/Course Material EXPERIMENTAL/02. Statistical Maps II/Readings and Practice/Fast Food.csv",
                      col_types = cols(Class = col_factor(levels = c("Chicken", "Hamburger", "Pizza", "Sub"))))

```

I can map this using ggmap. First, retrieve base map:
```{r}
Toronto_bm <- get_map("Bloor and Yonge, Toronto", zoom = 10) 
```

Check the base map
```{r}
ggmap(Toronto_bm)
```

Map the fast food establishments:
```{r}
ggmap(Toronto_bm) + geom_point(data = Fast_Food, aes(x = LONGITUDE, y = LATITUDE, shape = Class))
```

Remove observations that are north of 44.00 and west of -80.00
```{r}
Fast_Food <- Fast_Food[Fast_Food$LONGITUDE >= -79.8,]
Fast_Food <- Fast_Food[Fast_Food$LATITUDE <= 43.85,]
```

Redo the map  of the fast food establishments:
```{r}
ggmap(Toronto_bm) + geom_point(data = Fast_Food, aes(x = LONGITUDE, y = LATITUDE, shape = Class))
```

Then I saved to RData format.
```{r}
save(Fast_Food, file = "Fast_Food.RData")
```

