---
title: "Activity: Area Data V"
output: html_notebook
---

# Activity: Area Data V

## Practice questions

Answer the following questions:

1. Explain the main assumptions for linear regression models.
2. How is Moran's I used as a diagnostic in regression analysis?
3. Residual spatial autocorrelation is symptomatic of what issues in regression analysis?
4. What does it mean for a model to be linear in the coefficients?
5. What is the purpose of transforming variables for regression analysis?

## Learning objectives

In this activity, you will:

1. Explore a spatial dataset.
2. Conduct linear regression analysis.
3. Conduct diagnostics for residual spatial autocorrelation.
4. Propose ways to improve your analysis.

## Suggested reading

O'Sullivan D and Unwin D (2010) Geographic Information Analysis, 2nd Edition, Chapter 7. John Wiley & Sons: New Jersey. 

## Preliminaries

For this activity you will need the following:

* An R markdown notebook version of this document (the source file).

* A package called `geog4ga3`.

It is good practice to clear the working space to make sure that you do not have extraneous items there when you begin your work. The command in R to clear the workspace is `rm` (for "remove"), followed by a list of items to be removed. To clear the workspace from _all_ objects, do the following:
```{r}
rm(list = ls())
```

Note that `ls()` lists all objects currently on the worspace.

Load the libraries you will use in this activity. In addition to `tidyverse`, you will need `sf` and `geog4ga3`:
```{r}
library(tidyverse)
library(sf)
library(spdep)
library(geog4ga3)
```

Begin by loading the data files you will use in this activity:
```{r}
data("HamiltonDAs")
data("trips_by_mode")
data("travel_time_car")
```

`HamiltonDAs` are the Dissemination Areas for Hamilton CMA, which coincide with the Traffic Analysis Zones (TAZ) of the Transportation Tomorrow Survey of 2011. The dataframe `trips_by_mode`
includes the number of trips by mode of transportation by TAZ (equivalently DA), as well as other useful information from the 2011 census for Hamilton CMA. Finally, the dataframe `travel_time_car`
includes the travel distance/time from TAZ/DA centroids to Jackson Square in downtown Hamilton.

The data for this activity were retrieved from the 2011 Transportation Tomorrow Survey [TTS](http://www.transportationtomorrow.on.ca/), the periodic travel survey of the Greater Toronto and Hamilton Area, as well as data from the 2011 Canadian Census [Census Program](http://www12.statcan.gc.ca/census-recensement/index-eng.cfm).

Before beginning the activity, join the information on trips and travel time to the `sf` object. Note that to complete the join, the identifier (in this case `GTA06`) must be in the same format in both data frames:
```{r}
travel_time_car$GTA06 <- factor(travel_time_car$GTA06)

# Travel time
HamiltonDAs <- left_join(HamiltonDAs, travel_time_car, by = "GTA06")
# Trips by mode
HamiltonDAs <- left_join(HamiltonDAs, trips_by_mode, by = "GTA06")
```

The analysis will be based on travel by car in the Hamilton CMA. Calculate the proportion of trips by car by TAZ:
```{r}
HamiltonDAs <- mutate(HamiltonDAs, Auto_driver.prop = Auto_driver / (Auto_driver + Cycle + Walk))
```

Note that the proportion of people who travelled by car as passengers are not included in the denominator of the proportion! This is because every trip as a passenger is already included in trips with one driver.

## Activity

1. Examine your dataframe. What variables are included? Are there any missing values?

```{r}
summary(HamiltonDAs)
```


2. Map the variable `Auto_driver.prop`, and use Moran's I to test for spatial autocorrelation. What do you find? Is autocorrelation in this variable a sign that autocorrelation will be an issue in regression analysis?

```{r}
ggplot(data = HamiltonDAs) + 
  geom_sf(aes(fill = Auto_driver.prop)) +
  scale_fill_distiller(direction = 1)
```

Create listw
```{r}
HamiltonDAs.w <- nb2listw(poly2nb(pl = as(HamiltonDAs, "Spatial")))
```

Moran's I:
```{r}
moran.test(HamiltonDAs$Auto_driver.prop, HamiltonDAs.w)
```

```{r}
localmoran.map(HamiltonDAs, HamiltonDAs.w, "Auto_driver.prop", by = "GTA06")
```



3. Estimate regression model using the variables `Pop_Density` and travel time in `minutes`. Discuss this model.

```{r}
model1 <- lm(formula = Pop_Density ~ minutes, data = HamiltonDAs)
summary (model1)

ggplot(data = HamiltonDAs, aes(x = minutes, y = Pop_Density))+
  geom_point() +
  stat_function(fun = function(x)(5441.82 - 187.19 * x), 
                geom ="line", color = "blue", size = 1) +
  geom_vline(xintercept = 0) + geom_hline(yintercept = 0) 
#The values 5441.82 and 187.19 are taken from the summary of model 1 under coefficients. This is one exmaple of a model, other models can also be made depending on the function that you want to fit to the data. 
```

4. Examine residuals of the model. Are they spatially independent?
```{r}
HamiltonDAs <- mutate(HamiltonDAs, Pop_Density.log = log(Pop_Density))
model2 <- lm(formula = Pop_Density.log ~ minutes, data = HamiltonDAs)
summary (model2)
```
```{r}
moran.test(model1$residuals, HamiltonDAs.w)
```

```{r}
moran.test(model2$residuals, HamiltonDAs.w)
```



5. Propose ways to improve your model.