---
title: "18 Session 18: Spatially Continuous Data IV"
output: html_notebook
---

#Practice questions

Answer the following questions:

1. What does "Best" in BLUP mean?
2. What is the advantage of kriging over other interpolation approaches?
3. How is the the autocovariance used to produce optimal predictions?

#Learning objectives

In this activity, you will:

1. Use kriging to interpolate a 

#Suggested reading

O'Sullivan D and Unwin D (2010) Geographic Information Analysis, 2nd Edition, Chapters 8 and 9. John Wiley & Sons: New Jersey.

#Preliminaries

For this activity you will need the following:

* This R markdown notebook.

* A file called "Wolfcamp Aquifer.RData"

The data is a set of piezometric head (watertable pressure) observations of the Wolfcamp Aquifer in Texas (https://en.wikipedia.org/wiki/Hydraulic_head). Measures of pressure can be used to infer the flow of underground water, since water flows from high to low pressure areas.

These data were collected to evaluate potential flow of contamination related to a high level toxic waste repository in Texas. The Deaf Smith county site in Texas was one of three potential sites proposed for this repository. Beneath the site is a deep brine aquifer known as the Wolfcamp aquifer that may serve as a conduit of contamination leaking from the repository.

The data set consists of 85 georeferenced measurements of piezometric head. Possible applications of interpolation are to determine sites at risk and to quantify uncertainty of the interpolated surface, to evaluate the best locations for monitoring stations.

It is good practice to clear the working space to make sure that you do not have extraneous items there when you begin your work. The command in R to clear the workspace is `rm` (for "remove"), followed by a list of items to be removed. To clear the workspace from _all_ objects, do the following:
```{r}
rm(list = ls())
```

Note that `ls()` lists all objects currently on the worspace.

Load the libraries you will use in this activity (load other packages as appropriate). 
```{r}
library(tidyverse)
library(gstat)
library(spdep)
```

Load dataset:
```{r}
load("Wolfcamp Aquifer.RData")
```

Convert to a `SpatialPointsDataFrame`:
```{r}
aquifer.sp <- aquifer
coordinates(aquifer.sp) <- ~X+Y
```

#Activity

1. Partner with a fellow student to analyze the dataset provided.

2. Use kriging to estimate the underlying field. Justify your modeling choices.

3. Discuss your results.

4. Imagine that you had to compare different modeling approaches (e.g., kriging, IDW). Propose a protocol to decide which method is more accurate. 