---
title: "07 Activity 7: Point Pattern Analysis III"
output: html_notebook
---

#Practice questions

Answer the following questions:

1. List and explain two limitations of quadrat analysis.
2. What is clustering? What could explain a clustering in a set of events?
3. What is regularity? What could explain it?
4. Describe the concept of nearest neighbors.
5. What is a cumulative distribution function?

#Learning objectives

In this activity, you will:

1. Explore a dataset using distance-based approaches.
2. Compare the characteristics of different types of patterns.
3. Discuss ways to evaluate how confident you are that a pattern is random.

#Suggested reading

O'Sullivan D and Unwin D (2010) Geographic Information Analysis, 2nd Edition, Chapter 5. John Wiley & Sons: New Jersey.

#Preliminaries

For this activity you will need the following:

* This R markdown notebook.

* A dataset called `Toronto Business Points.RData`.

It is good practice to clear the working space to make sure that you do not have extraneous items there when you begin your work. The command in R to clear the workspace is `rm` (for "remove"), followed by a list of items to be removed. To clear the workspace from _all_ objects, do the following:
```{r}
rm(list = ls())
```

Note that `ls()` lists all objects currently on the worspace.

Load the libraries you will use in this activity. In addition to `tidyverse`, you will need `spatstat`, a package designed for the analysis of point patterns (you can learn about `spatstat` [here](https://cran.r-project.org/web/packages/spatstat/vignettes/getstart.pdf) and [here](http://spatstat.org/resources/spatstatJSSpaper.pdf)):
```{r}
library(tidyverse)
library(spatstat)
```

In the practice that preceded this activity, you learned about the concepts of intensity and density, about quadrats, and also how to create density maps. 

Begin by loading the data that you will use in this activity:
```{r}
load("Toronto Business Points.RData")
```

If you inspect your workspace, you will see that the following `ppp` objects are there:

* `Fast_Food.ppp`
* `Gas_Stands.ppp`
* `Paez_Mart.ppp`

These are locations of fast food restaurants and gas stands in Toronto (data are from 2008). Paez Mart on the other hand is my project to cover Toronto with convenience stores. The points are the planned locations of the stores. 

Please note that the exact same data are provided in dataframe formand that `Toronto` is the city boundary. You do not need to use these dataframes for the present activity.

You can check the contents of `ppp` objects by means of `summary`:
```{r}
summary(Fast_Food.ppp)
```

#Activity

1. Calculate the event-to-event distances to nearest neighbors using the function `nndist()`. Do this for all fast food establishments (pooled) and then for each type of establishment (i.e, "Chicken", "Hamburger", "Pizza", "Sub").

2. Create Stienen diagrams using the distance vectors obtained in Step 1. Discuss the diagrams with a fellow student.

3. Plot the empirical G-function for all fast food establishments (pooled) and then for each type of establishment (i.e, "Chicken", "Hamburger", "Pizza", "Sub").

4. Is there evidence of clustering/regularity? 

5. How confident are you to make a decision whether the patterns are not random? What could you do to assess your confidence in making a decision whether the patterns are random? Explain.