---
title: "Test Ideas Module 3"
output: html_notebook
---


```{r}
library(sp)
library(spdep)
library(rgdal)
library(adespatial)
library(tmap)
```
  
```{r}
shape <- readOGR(dsn=".", layer = "WiFi_Collector_Wicked_Free_")
```

Find parameters for the simulation
```{r}
N <- length(shape) #number of events
w <- bbox(shape) #bounding box
```

Obtain coordinates on a regular grid:
```{r}
ng <- 25 # number of grid elements
x <- seq(min(w[1,]), to = max(w[1,]), by = (max(w[1,]) - min(w[1,]))/ng) 
y <- seq(min(w[2,]), to = max(w[2,]), by = (max(w[2,]) - min(w[2,]))/ng)
```

Use these vectors of coordinates to create the points for a regular grid:
```{r}
pts = sp::SpatialPoints(expand.grid(x = x, y = y))
proj4string(pts) <- proj4string(shape)
```

Plot these points:
```{r}
plot(pts)
```

To obtain the Moran Eigenvector Maps we must first obtain the contiguities for the system.

Convert points (`pts`) to nearest neighbors using the `knearneigh` function, selecting k=4 (number of neighbors):
```{r}
nn <- knn2nb(knearneigh(pts, k = 4, longlat = TRUE))
```

Spatial weights
```{r}
spw <- nb2listw(nn)
```

Now the spatial weights can be used to obtain Moran eigenvector maps using `adespatial::mem`:
```{r}
MEM <- mem(spw)
```

Add MEM to points:
```{r}
pts.mem <- SpatialPointsDataFrame(pts, data.frame(MEM))
```

```{r}
tm_shape(pts.mem) + tm_dots("MEM50", size = 0.5)
```