---
title: "19-Area-Data-examples"
output: html_notebook
---

Examples for slides. 

```{r message=FALSE, warning=FALSE}
library(geog4ga3)
library(patchwork)
library(tidyverse)
library(spdep)
library(sf)
#library(gridExtra)
```

Read the data used in this chapter. This is an object of class `sf` (simple feature) with the census tracts of Hamilton CMA and some selected population variables from the 2011 Census of Canada:
```{r}
data(Hamilton_CT)
```

You can quickly verify the contents of the dataframe by means of `summary`:
```{r}
summary(Hamilton_CT)
```

## Spatial Moving Averages and Simulation

In the preceding chapter and activity you learned about different criteria to define proximity for the analysis of area data, and how spatial weights matrices can be used to code patterns of proximity among zones in a spatial system. Furthermore, you also saw how spatial weights matrices can be used to calculate spatial moving averages, which in turn can be used to explore spatial patterns in area data.

We will begin this chapter by briefly revisiting some of these notions. In the following chunk, we create a spatial weights matrix for Hamilton CMA census tracts based on the adjacency criterion:
```{r}
# Function `poly2nb()` builds a list of neighbors based on contiguous boundaries. The argument for this function is an object of class "sf", which contains multi-polygon objects. 

# Function `nb2listw()` takes a list of neighbors and creates a matrix of spatial weights in the form of a list. Together, these two functions create a spatial weights matrix for the Census Tracts in Hamilton.

Hamilton_CT.nb <- poly2nb(pl = Hamilton_CT)
Hamilton_CT.w <- nb2listw(Hamilton_CT.nb)
```

Once that you have a matrix of spatial weights, it can be used to calculate the spatial moving average. In this example, we calculate the spatial moving average of the variable for population density, i.e., `POP_DENSITY` which is found in the `sf` dataframe:
```{r}
# The function `lag.listw()` takes as argument the population density by census tracts in Hamilton, and calculates the moving average, with the "moving" part given by the local neighborhoods around each zone as defined by `Hamilton_CT.w`

POP_DENSITY.sma <- lag.listw(Hamilton_CT.w, Hamilton_CT$POP_DENSITY)
```

After calculating the spatial moving average of population density, we can join this new variable to both the `sf` and `SpatialPolygonsDataFrame` objects:
```{r}
Hamilton_CT$POP_DENSITY.sma <- POP_DENSITY.sma
```

As you saw in your last activity, the spatial moving average can be used in two ways to explore the spatial pattern of an area variable: as a smoother and by means of a scatterplot, combined with the original variable.

## The Spatial Moving Average as a Smoother

The spatial moving average, when mapped, is essentially a smoothing technique. What do we mean by smoothing? By reporting the average of the neighbors instead of the actually observed value of the variable, we reduce the amount of variability that is communicated. This often can make it easier to distinguish the overall pattern, at the cost of some information loss (think of how when mapping quadrats we lost some information/detail by calculating the intensity for areas).

We can illustrate the use of the spatial moving average as a smoother with the help of a little simulation.

To simulate a random spatial variable, we can randomize the observations that we already have, reassigning them at random to areas in the system. This is accomplished as follows:
```{r}
# By sampling at random and without replacement from the original variable, we create a null landscape. We will call this `POP_DENSITY_s1`, where the "s1" part is to indicate that this is our first simulated random landscape. We will actually repeat this process below.

POP_DENSITY_s1 <- sample(Hamilton_CT$POP_DENSITY)
```

Calculate the spatial moving average for this randomized variable (i.e., null landscape):
```{r}
# We use the function `lag.listw()` to calculate the spatial moving average, but now for the null landscape we just simulated. 

POP_DENSITY_s1.sma <- lag.listw(Hamilton_CT.w, POP_DENSITY_s1)
```

Once that you have seen how to randomize the variable, repeat the process to simulate a total of eight new variables/null landscapes, and calculate their spatial moving averages:
```{r}
# Note that we are creating 8 null landscapes based on our original population density variable, and that we are calculating the spatial moving average for each of them. Each simulation has a new name: s2, s3, s4,..., s8. 

# Null landscape/simulation #2
POP_DENSITY_s2 <- sample(Hamilton_CT$POP_DENSITY)
POP_DENSITY_s2.sma <- lag.listw(Hamilton_CT.w, POP_DENSITY_s2)

# Null landscape/simulation #3
POP_DENSITY_s3 <- sample(Hamilton_CT$POP_DENSITY)
POP_DENSITY_s3.sma <- lag.listw(Hamilton_CT.w, POP_DENSITY_s3)

# Null landscape/simulation #4
POP_DENSITY_s4 <- sample(Hamilton_CT$POP_DENSITY)
POP_DENSITY_s4.sma <- lag.listw(Hamilton_CT.w, POP_DENSITY_s4)

# Null landscape/simulation #5
POP_DENSITY_s5 <- sample(Hamilton_CT$POP_DENSITY)
POP_DENSITY_s5.sma <- lag.listw(Hamilton_CT.w, POP_DENSITY_s5)

# Null landscape/simulation #6
POP_DENSITY_s6 <- sample(Hamilton_CT$POP_DENSITY)
POP_DENSITY_s6.sma <- lag.listw(Hamilton_CT.w, POP_DENSITY_s6)

# Null landscape/simulation #7
POP_DENSITY_s7 <- sample(Hamilton_CT$POP_DENSITY)
POP_DENSITY_s7.sma <- lag.listw(Hamilton_CT.w, POP_DENSITY_s7)

# Null landscape/simulation #8
POP_DENSITY_s8 <- sample(Hamilton_CT$POP_DENSITY)
POP_DENSITY_s8.sma <- lag.listw(Hamilton_CT.w, POP_DENSITY_s8)
```

Next, we will add all the null landscapes that you just simulated to the dataframes, as well as their spatial moving averages. This is useful for mapping and plotting purposes:
```{r}
# Here we add the simulated landscapes to the `sf` dataframe.
Hamilton_CT$POP_DENSITY_s1 <- POP_DENSITY_s1
Hamilton_CT$POP_DENSITY_s2 <- POP_DENSITY_s2
Hamilton_CT$POP_DENSITY_s3 <- POP_DENSITY_s3
Hamilton_CT$POP_DENSITY_s4 <- POP_DENSITY_s4
Hamilton_CT$POP_DENSITY_s5 <- POP_DENSITY_s5
Hamilton_CT$POP_DENSITY_s6 <- POP_DENSITY_s6
Hamilton_CT$POP_DENSITY_s7 <- POP_DENSITY_s7
Hamilton_CT$POP_DENSITY_s8 <- POP_DENSITY_s8

# Here we add the spatial moving averages of the simulated landscapes to the `sf` dataframe.
Hamilton_CT$POP_DENSITY_s1.sma <- POP_DENSITY_s1.sma
Hamilton_CT$POP_DENSITY_s2.sma <- POP_DENSITY_s2.sma
Hamilton_CT$POP_DENSITY_s3.sma <- POP_DENSITY_s3.sma
Hamilton_CT$POP_DENSITY_s4.sma <- POP_DENSITY_s4.sma
Hamilton_CT$POP_DENSITY_s5.sma <- POP_DENSITY_s5.sma
Hamilton_CT$POP_DENSITY_s6.sma <- POP_DENSITY_s6.sma
Hamilton_CT$POP_DENSITY_s7.sma <- POP_DENSITY_s7.sma
Hamilton_CT$POP_DENSITY_s8.sma <- POP_DENSITY_s8.sma
```

It would be useful to compare the original landscape of population density to the null landscapes that you created before. To create a single figure with choropleth maps of the empirical variable and the eight simulated variables using the `facet_wrap()` function of `ggplot2`, we must first reorganize the data so that all the population density variables are in a single column, and all spatial moving average variables are also in a single column. Further, we need a new column to identifies which variable the values in this column correspond to. We will solve this little data management problem by copying only the data we are interested in into a new dataframe (by means of `select()`), and then _gathering_ the spatial moving averages into a single column: 
```{r}
#"Hamilton_CT2 is a new dataframe. Here, the pipe operators (%>%) are used to pass the original dataframe to the select() function, and then the output of that is passed on to the `gather()` function. Notice that we are selecting the empirical spatial moving average and the 8 simulated instances of population densities. 

Hamilton_CT2 <- Hamilton_CT %>% # This pipe operator passes the dataframe to `select()`
  # `select()` keeps only the spatial moving averages and geometry
  select(POP_DENSITY.sma, 
         POP_DENSITY_s1.sma,
         POP_DENSITY_s2.sma,
         POP_DENSITY_s3.sma,
         POP_DENSITY_s4.sma,
         POP_DENSITY_s5.sma,
         POP_DENSITY_s6.sma,
         POP_DENSITY_s7.sma,
         POP_DENSITY_s8.sma,
         geometry) %>% # This pipe operator passes the dataframe with only the spatial moving average variables and the geometry to `gather()`
  # `gather()` places all variables with the exception of `geometry` in a single column named `DENSITY_SMA` and creates a new variable called `VAR` with the names of the original columns (i.e., POP_DENSITY.sma, POP_DENSITY_s1.sma, etc.)
  pivot_longer(cols = -geometry, names_to = "VAR", values_to = "DENSITY_SMA") %>% st_as_sf()
```

Now the new dataframe with all spatial moving averages in a single column can be used to create choropleth maps. The function `facet_wrap()` is used to create facet plots so that we can place all maps in a single figure:
```{r}
ggplot() + 
  geom_sf(data = Hamilton_CT2, 
          aes(fill = DENSITY_SMA), color = NA) + 
  facet_wrap(~VAR, ncol = 3) + # We are creating multiple plots for single data frame by means of the "facet_wrap" function.
  scale_fill_distiller(palette = "YlOrRd", direction = 1) + # Select palette for colors 
  labs(fill = "Pop Den SMA") + # Change the label of the legend
  theme(axis.text.x = element_blank(), 
        axis.text.y = element_blank()) # Remove the axis labels to avoid cluttering the plots
```

The empirical variable is the map in the upper left corner (labeled `POP_DENSITY.sma`). The remaining 8 maps are simulated variables. Would you say the map of the empirical variable is fairly different from the map of the simulated variables? What are the key differences?

An additional advantage of the spatial moving average is its use in the development of scatterplots. The information below provides further examples of exploring spatial moving averages with scatterplots.

## Spatial Moving Average Scatterplots

Let us explore the use of spatial moving average scatterplots. First, we will extract the density information from the original `sf` object, reorganize, and bind to `Hamilton_CT2` so that we can plot using faceting:
```{r}
Hamilton_CT2 <- Hamilton_CT2 %>% # Pass `Hamilton_CT2` as the first argument of `data.frame()`
  data.frame(Hamilton_CT %>% # Pass `Hamilton_CT` to `st_drop_geometry()`
               st_drop_geometry() %>% # Drop the geometry because it is already available in `Hamilton_CT2`.
               # Select from `Hamilton_CT` the original population density and the 8 null landscapes simulated from it.
               select(POP_DENSITY,
                      POP_DENSITY_s1,
                      POP_DENSITY_s2,
                      POP_DENSITY_s3,
                      POP_DENSITY_s4,
                      POP_DENSITY_s5,
                      POP_DENSITY_s6,
                      POP_DENSITY_s7,
                      POP_DENSITY_s8) %>% # Pass the result to `gather()`  
               pivot_longer(cols = everything(), names_to = "VAR", values_to = "DENSITY") %>% # Copy all density variables to a single column, and create a new variable called `VAR` with the names of the original columns (i.e., POP_DENSITY, POP_DENSITY_s1, etc.) 
               select(DENSITY)) # Drop VAR from the the dataframe
```

After reorganizing the data we can create the scatterplot of the empirical population density and its spatial moving average, as well as the scatterplots of the simulated variables and their spatial moving averages for comparison (the plots include the 45 degree line). Again, the use of `facet_wrap()` allows us to put all plots in a single figure:
```{r}
#We are adding a geom and line (slope = 1)
ggplot(data = Hamilton_CT2, aes(x = DENSITY, y = DENSITY_SMA, color = VAR)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm") +
  geom_abline(slope = 1, 
              intercept = 0, color = "white") +
  coord_equal() +
  facet_wrap(~ VAR, ncol = 3) +
  theme(plot.background = element_rect(fill = "black"),
        legend.background = element_rect(fill = "black"),
        legend.key = element_rect(fill = "black"),
        legend.text = element_text(color = "white"),
        legend.title = element_text(color = "white"),
        legend.position = "none",
        axis.text = element_blank(),
        panel.background = element_rect(fill = "black"),
        strip.background = element_rect(fill = "black"),
        strip.text = element_text(color = "white"),
        panel.grid = element_blank())

ggsave("simulated-scatterplots.png")
```

Calculate the mean-centered version of POP_DENSIT, and then its spatial moving average:
```{r}
df_mean_center_scatterplot <- transmute(Hamilton_CT, # Modify values in dataframe
                                        Density_z = POP_DENSITY - mean(POP_DENSITY), # Subtract the mean, so that the variable now is deviations from the mean 
                                        SMA_z = lag.listw(Hamilton_CT.w, Density_z)) # Calculate the spatial moving average of the newly created variable `Density_z`
```

Compare the following two plots. You will see that they are identical, but in the mean-centered one the origin of the axes coincides with the means of $x$ and the spatial moving average of $x$. In other words, we have the same data, but we have displaced the origin of the plot:
```{r}
# Create a scatterplot of population density and its spatial moving average
sc1 <- ggplot(data = filter(Hamilton_CT2, VAR == "POP_DENSITY.sma"),
              aes(x = DENSITY, y = DENSITY_SMA)) +
  geom_point(color = "white", alpha = 0.2) +
  geom_abline(slope = 1, intercept = 0, color = "white", linetype = "dashed") +
  geom_smooth(method = "lm", color = "white") +
  ggtitle("Population Density") +
  coord_equal() +
  theme(text = element_text(color = "white"),
        title = element_text(size = 8),
        axis.text.x = element_text(color = "white", size = 6),
        axis.text.y = element_text(color = "white", size = 6),
        plot.background = element_rect(fill = "black"),
        legend.background = element_rect(fill = "black"),
        legend.key = element_rect(fill = "black"),
        legend.text = element_text(color = "white"),
        legend.title = element_text(color = "white"),
        legend.position = "none",
        axis.text = element_blank(),
        panel.background = element_rect(fill = "black"),
        strip.background = element_rect(fill = "black"),
        strip.text = element_text(color = "white"),
        panel.grid = element_blank())

# Create a scatterplot of the mean-centered population density, and its spatial moving average
sc2 <- ggplot(data = df_mean_center_scatterplot, 
              aes(x = Density_z, y = SMA_z)) +
  geom_point(color = "white", alpha = 0.2) +
  geom_abline(slope = 1, color = "white", intercept = 0, linetype = "dashed") +
  geom_smooth(method = "lm", formula = y ~ x-1, color = "white") +
  ggtitle("Mean-Centered Population Density") +
  coord_equal() +
  theme(text = element_text(color = "white"),
        title = element_text(size = 8),
        axis.text.x = element_text(color = "white", size = 6),
        axis.text.y = element_text(color = "white", size = 6),
        plot.background = element_rect(fill = "black"),
        legend.background = element_rect(fill = "black"),
        legend.key = element_rect(fill = "black"),
        legend.text = element_text(color = "white"),
        legend.title = element_text(color = "white"),
        legend.position = "none",
        axis.text = element_blank(),
        panel.background = element_rect(fill = "black"),
        strip.background = element_rect(fill = "black"),
        strip.text = element_text(color = "white"),
        panel.grid = element_blank()
        )

# Use `grid.arrange()` to place the two plots in a single figure 
sc1 + sc2# +
  #plot_layout(nrow = 1, widths = c(1, 1)) &
  #theme(plot.margin = unit(c(.2,.2,.2,.2), "cm")) &
  #plot_annotation(theme = theme(plot.background = element_rect(color = "black", fill  = 'black')))
ggsave("mean-centered-variable.png")
```

```{r}
ggplot(data = df_mean_center_scatterplot, 
       aes(x = Density_z, y = SMA_z)) +
  geom_point(color = "white", alpha = 0.2) +
  geom_hline(yintercept = 0, color = "white") +
  geom_vline(xintercept = 0, color = "white") +
  # You can also add annotations to plots by using `annotate()`. The inputs are the kind of annotation; in this case "text", but it could be circles, arrows, rectangles, labels, and other things. For text, you need a label, and coordinates for the annotation.
  annotate("text", label = "Q1: Positive", x= 2000, y = 2500, color = "white") +
  annotate("text", label = "Q4: Negative", x= 2000, y = -2500, color = "white") +
  annotate("text", label = "Q2: Negative", x= -2000, y = 2500, color = "white") +
  annotate("text", label = "Q3: Positive", x= -2000, y = -2500, color = "white") +
  coord_equal() +
  theme(text = element_text(color = "white"),
        title = element_text(size = 8),
        axis.text.x = element_text(color = "white"),
        axis.text.y = element_text(color = "white"),
        plot.background = element_rect(fill = "black"),
        legend.background = element_rect(fill = "black"),
        legend.key = element_rect(fill = "black"),
        legend.text = element_text(color = "white"),
        legend.title = element_text(color = "white"),
        legend.position = "none",
        axis.text = element_blank(),
        panel.background = element_rect(fill = "black"),
        strip.background = element_rect(fill = "black"),
        strip.text = element_text(color = "white"),
        panel.grid = element_blank()
        )
ggsave("quadrants-scatterplot.png")
```