---
title: "Activity 10: Area Data I"
output: html_notebook
---

# Activity 10: Area Data I

Remember, you can download the source file for this activity from [here](https://github.com/paezha/Spatial-Statistics-Course).

## Practice questions

Answer the following questions:

1. What is a key difference between area data and point data?
2. What is a choropleth map?
3. What is a cartogram?
4. What are the advantages and disadvantages of these mapping techniques?

## Learning objectives

In this activity, you will:

1. Create choroplet maps using census data.
2. Think about possible underlying process that could explain the pattern.
3. Think about ways to decide whether a landscape is random when working with area data.

## Suggested reading

O'Sullivan D and Unwin D (2010) Geographic Information Analysis, 2nd Edition, Chapter 7. John Wiley & Sons: New Jersey.

## Preliminaries

For this activity you will need the following:

* An R markdown notebook version of this document (the source file).

* A package called `geog4ga3`.

It is good practice to clear the working space to make sure that you do not have extraneous items there when you begin your work. The command in R to clear the workspace is `rm` (for "remove"), followed by a list of items to be removed. To clear the workspace from _all_ objects, do the following:
```{r}
rm(list = ls())
```

Note that `ls()` lists all objects currently on the worspace.

Load the libraries you will use in this activity. In addition to `tidyverse`, you will need `spatstat`, a package designed for the analysis of point patterns (you can learn about `spatstat` [here](https://cran.r-project.org/web/packages/spatstat/vignettes/getstart.pdf) and [here](http://spatstat.org/resources/spatstatJSSpaper.pdf)):
```{r}
library(tidyverse)
library(rgdal)
library(broom)
library(cartogram)
```

In the practice that preceded this activity, you learned about the area data and visualization techniques for area data.

Begin by loading the data that you will use in this activity:
```{r}
Hamilton_CT <- readOGR(".", layer = "Hamilton CMA CT", integer64 = "allow.loss")
```

You can obtain new (calculated) variables as follows. For instance, to obtain the proportion of residents who are between 20 and 34 years old, and between 35 and 49:
```{r}
Hamilton_CT@data <- mutate(Hamilton_CT@data, Prop20to34 = (AGE_20_TO_ + AGE_25_TO_ + AGE_30_TO_)/POPULATION, Prop35to49 = (AGE_35_TO_ + AGE_40_TO_ + AGE_45_TO_)/POPULATION)
```

This is a `SpatialPolygonDataFrame`. Convert to a dataframe ("tidy" it) for plotting using `ggplot2`:
```{r}
Hamilton_CT.t <- tidy(Hamilton_CT, region = "TRACT")
Hamilton_CT.t <- rename(Hamilton_CT.t, TRACT = id)
```

Rejoin the data:
```{r}
Hamilton_CT.t <- left_join(Hamilton_CT.t, Hamilton_CT@data, by = "TRACT")
```

#Activity

1. Create choropleth maps for the proportion of the population who are 20 to 34 years old, 35 to 49 years old, 50 to 65 years old, and 65 and older. 

2. Show your maps to a fellow student. What patterns do you notice in the distribution of age in Hamilton?

3. Devise a rule to decide whether the pattern observed in a choropleth map is random.