---
title: "Data Preprocessing: Point Pattern Analysis I"
output: html_notebook
---

#Introduction

This activity will introduce more formally the concept of point patterns and analysis. 

My initial idea is to implement 

Clear workspace:
```{r}
rm(list = ls())
```

Load libraries:
```{r}
library(tidyverse)
library(spdep)
library(adespatial)
```

Load the point patterns that I created in Session 2:
```{r}
load("Data2.RData")
```

Select dataset to plot:
```{r}
data = PointPattern2
```

Plot (use `theme` to remove gridlines):
```{r}
ggplot() + 
  geom_point(data = data, aes(x = x, y = y)) + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  coord_fixed()
```

Plot (use `theme` to remove gridlines):
```{r}
ggplot() + 
  geom_point(data = data, aes(x = x, y = y)) + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  geom_vline(xintercept = seq(from = 0, to = 1, by = 1)) +
  geom_hline(yintercept = seq(from = 0, to = 1, by = 1)) +
  coord_fixed()
```

Note that `geom_bind2` can be used to produce _density maps_.
```{r}
ggplot() + 
  geom_vline(xintercept = seq(from = 0, to = 1, by = 1)) +
  geom_hline(yintercept = seq(from = 0, to = 1, by = 1)) +
  geom_bin2d(data = data, aes(x = x, y = y), binwidth = c(1, 1)) +
  geom_point(data = data, aes(x = x, y = y), color = "white", shape = 16, size = 2) + 
  coord_fixed()
```

Note that `geom_bind2` can be used to produce _density maps_.
```{r}
q_size = 0.25
ggplot() + 
  geom_bin2d(data = data, aes(x = x, y = y), binwidth = c(q_size, q_size)) +
  geom_vline(xintercept = seq(from = 0, to = 1, by = q_size)) +
  geom_hline(yintercept = seq(from = 0, to = 1, by = q_size)) +
  geom_point(data = data, aes(x = x, y = y), color = "white", shape = 16, size = 2) +
  coord_fixed()
```

Explore the different patterns.
```{r}
data = PointPattern1
q_size = 0.25
ggplot() + 
  geom_bin2d(data = data, aes(x = x, y = y), binwidth = c(q_size, q_size)) +
  geom_vline(xintercept = seq(from = 0, to = 1, by = q_size)) +
  geom_hline(yintercept = seq(from = 0, to = 1, by = q_size)) +
  geom_point(data = data, aes(x = x, y = y), color = "white", shape = 16, size = 2) +
  coord_fixed()
```

Put all patterns in a single dataframe:
```{r}
PointPatterns <- rbind(PointPattern1, PointPattern2, PointPattern3)
```

Add a regular pattern:
```{r}
junk <- expand.grid(x = seq(from = 0.075, to = 0.825, by = 0.15), 
            y = seq(from = 0.05, to = 0.85, by = 0.20))
junk2 <- expand.grid(x = seq(from = 0.15, to = 0.90, by = 0.15), 
            y = seq(from = 0.15, to = 0.95, by = 0.20))
PointPattern4 <- rbind(junk, junk2)
ggplot() + geom_point(data = PointPattern4, aes(x = x, y = y)) + 
  coord_fixed()
```

```{r}
PointPatterns <- rbind(PointPatterns, PointPattern4)
```

Add factors to identify the patterns:
```{r}
PointPatterns$Pattern <- factor(rep(1:4, each = 60), 
                                levels = c("1", "2", "3", "4"),
                                labels = c("Pattern 1", "Pattern 2", "Pattern 3", "Pattern 4"))
```

Save:
```{r}
save(PointPatterns, file = "Data3.RData")
```

