---
title: "09 Session 9: Area Data I - Bonus Material"
output: html_notebook
---

#Preliminaries

To work with this bonus material you will need:

* This R markdown notebook.
* A dataset called `Toronto Business Points.RData`.

It is good practice to clear the working space to make sure that you do not have extraneous items there when you begin your work. The command in R to clear the workspace is `rm` (for "remove"), followed by a list of items to be removed. To clear the workspace from _all_ objects, do the following:
```{r}
rm(list = ls())
```

Note that `ls()` lists all objects currently on the worspace.

Load the libraries you will use in this activity. In addition to `tidyverse`, you will need `spatstat`, a package designed for the analysis of point patterns (you can learn about `spatstat` [here](https://cran.r-project.org/web/packages/spatstat/vignettes/getstart.pdf) and [here](http://spatstat.org/resources/spatstatJSSpaper.pdf)):
```{r}
library(tidyverse)
library(rgdal)
library(broom)
library(cartogram)
library(plotly)
```

In the practice that preceded this activity, you learned about the area data and visualization techniques for area data.

Begin by loading the data that you will use in this activity:
```{r}
Hamilton_CT <- readOGR(".", layer = "Hamilton CMA CT", integer64 = "allow.loss")
```

You can obtain new (calculated) variables as follows. For instance, to obtain the proportion of residents who are between 20 and 34 years old, and between 35 and 49:
```{r}
Hamilton_CT@data <- dplyr::transmute(Hamilton_CT@data, AREA = AREA, TRACT = TRACT,
                                     POPULATION = POPULATION,
                                     POP20to34 = (AGE_20_TO_ + AGE_25_TO_ + AGE_30_TO_),
                                     Prop20to34 = POP20to34/POPULATION, 
                                     POP35to49 = (AGE_35_TO_ + AGE_40_TO_ + AGE_45_TO_),
                                     Prop35to49 = POP35to49/POPULATION, 
                                     POP50to64 = (AGE_50_TO_ + AGE_55_TO_ + AGE_60_TO_),
                                     Prop50to64 = POP50to64/POPULATION,
                                     POP65Plus = (AGE_65_TO_ + AGE_70_TO_ + AGE_75_TO_ + AGE_80_TO_ + AGE_85),
                                     Prop65Plus = POP65Plus/POPULATION)
```

This is a `SpatialPolygonDataFrame`. Convert to a dataframe ("tidy" it) for plotting using `ggplot2`:
```{r}
Hamilton_CT.t <- tidy(Hamilton_CT, region = "TRACT")
Hamilton_CT.t <- rename(Hamilton_CT.t, TRACT = id)
```

Rejoin the data:
```{r}
Hamilton_CT.t <- left_join(Hamilton_CT.t, Hamilton_CT@data, by = "TRACT")
```

#Activity

1. Create choropleth maps for the proportion of the population who are 20 to 34 years old, 35 to 49 years old, 50 to 65 years old, and 65 and older. 

```{r}
ggplot() + 
  geom_polygon(data = Hamilton_CT.t, aes(x = long, y = lat, group = group, fill = cut_number(Prop20to34, 5)), color = "white") + 
  coord_fixed() +
  scale_fill_brewer(palette = "YlOrRd") +
  labs(fill = "Prop Age 20 to 34")

```

```{r}
ggplot() + 
  geom_polygon(data = Hamilton_CT.t, aes(x = long, y = lat, group = group, fill = cut_number(Prop35to49, 5)), color = "white") + 
  coord_fixed() +
  scale_fill_brewer(palette = "YlOrRd") +
  labs(fill = "Prop Age 35 to 49")

```

```{r}
ggplot() + 
  geom_polygon(data = Hamilton_CT.t, aes(x = long, y = lat, group = group, fill = cut_number(Prop50to64, 5)), color = "white") + 
  coord_fixed() +
  scale_fill_brewer(palette = "YlOrRd") +
  labs(fill = "Prop Age 50 to 64")

```

```{r}
ggplot() + 
  geom_polygon(data = Hamilton_CT.t, aes(x = long, y = lat, group = group, fill = cut_number(Prop65Plus, 5)), color = "white") + 
  coord_fixed() +
  scale_fill_brewer(palette = "YlOrRd") +
  labs(fill = "Prop Age 65+")

```

2. Show your maps to a fellow student. What patterns do you notice in the distribution of age in Hamilton?

3. Devise a rule to decide whether the pattern observed in a choropleth map is random.

# Bonus

Create cartograms of the same variables above. Since the cartograms are created based on the `SpatialPolygonsDataFrame`:
```{r}
Prop20to34.cartogram <- cartogram(shp = Hamilton_CT, weight = "POPULATION")
Prop35to49.cartogram <- cartogram(shp = Hamilton_CT, weight = "POPULATION")
Prop50to64.cartogram <- cartogram(shp = Hamilton_CT, weight = "POPULATION")
Prop65Plus.cartogram <- cartogram(shp = Hamilton_CT, weight = "POPULATION")
```

Tidy and restore the data:
```{r}
Prop20to34.cartogram.t <- tidy(Prop20to34.cartogram, region = "TRACT")
Prop20to34.cartogram.t <- rename(Prop20to34.cartogram.t, TRACT = id)
Prop20to34.cartogram.t <- left_join(Prop20to34.cartogram.t, select(Prop20to34.cartogram@data, TRACT, Prop20to34), by = "TRACT")
Prop20to34.cartogram.t <- rename(Prop20to34.cartogram.t, Proportion = Prop20to34)

Prop35to49.cartogram.t <- tidy(Prop35to49.cartogram, region = "TRACT")
Prop35to49.cartogram.t <- rename(Prop35to49.cartogram.t, TRACT = id)
Prop35to49.cartogram.t <- left_join(Prop35to49.cartogram.t, select(Prop35to49.cartogram@data, TRACT, Prop35to49), by = "TRACT")
Prop35to49.cartogram.t <- rename(Prop35to49.cartogram.t, Proportion = Prop35to49)

Prop50to64.cartogram.t <- tidy(Prop50to64.cartogram, region = "TRACT")
Prop50to64.cartogram.t <- rename(Prop50to64.cartogram.t, TRACT = id)
Prop50to64.cartogram.t <- left_join(Prop50to64.cartogram.t, select(Prop50to64.cartogram@data, TRACT, Prop50to64), by = "TRACT")
Prop50to64.cartogram.t <- rename(Prop50to64.cartogram.t, Proportion = Prop50to64)

Prop65Plus.cartogram.t <- tidy(Prop65Plus.cartogram, region = "TRACT")
Prop65Plus.cartogram.t <- rename(Prop65Plus.cartogram.t, TRACT = id)
Prop65Plus.cartogram.t <- left_join(Prop65Plus.cartogram.t, select(Prop65Plus.cartogram@data, TRACT, Prop65Plus), by = "TRACT")
Prop65Plus.cartogram.t <- rename(Prop65Plus.cartogram.t, Proportion = Prop65Plus)
```

Stack all cartograms into a single dataframe, add factor for frames:
```{r}
Cartograms <- rbind(data.frame(Prop20to34.cartogram.t, factor = "Age 20 to 34"),
                    data.frame(Prop35to49.cartogram.t, factor = "Age 35 to 49"),
                    data.frame(Prop50to64.cartogram.t, factor = "Age 50 to 64"),
                    data.frame(Prop65Plus.cartogram.t, factor = "Age 65+"))
```

Create plot with frames:
```{r}
Cartograms.plot <- ggplot(Cartograms, aes(x= long, y = lat, group = group, fill = cut_number(Proportion, 5))) + 
  geom_polygon(aes(frame = factor), color = "white") +
  scale_fill_brewer(palette = "YlOrRd") +
  coord_fixed() +
  theme(legend.position = "bottom") +
  labs(fill = "Proportion")
```

Plot:
```{r}
ggplotly(Cartograms.plot)
```

Try rectangular cartograms:
```{r}
library(recmap)
```

Rectangular algorithms require the centroids of the zones, which can be obtained by means of `sp::coordinates`
```{r}
xy_centroids <- coordinates(Hamilton_CT)
```

Create a frame for the rectangular cartograms:
```{r}
rec_frame <- data.frame(x = xy_centroids[,1], 
           y = xy_centroids[,2],
           dx = sqrt(Hamilton_CT$AREA) / 2 / (0.7 * 60 * cos(xy_centroids[,2] * pi / 180)),
           dy = sqrt(Hamilton_CT$AREA) / 2 / (0.7 * 60)) 
```

Generate the input for the `recmap` function. This is the frame for the cartograms `rec_frame` and a variable z:
```{r}
POP.rec <- data.frame(rec_frame, z = Hamilton_CT$POPULATION + 1, name = Hamilton_CT$TRACT)
head(POP.rec)
```

```{r}
POP.rec.cartogram <- recmap(POP.rec)
```

```{r}
plot(POP.rec.cartogram)
```

```{r}
ggplot(data = POP.rec.cartogram) + geom_rect(aes(xmin = x - dx, xmax = x + dx, ymin = y - dy, ymax = y + dy)) + coord_fixed()
```

```{r}
Cartograms.rec <- rbind(data.frame(POP.rec.cartogram,
                                   Proportion = Hamilton_CT$Prop20to34,
                                   factor = "Age 20 to 34"),
                        data.frame(POP.rec.cartogram,
                                   Proportion = Hamilton_CT$Prop35to49,
                                   factor = "Age 35 to 49"),
                        data.frame(POP.rec.cartogram,
                                   Proportion = Hamilton_CT$Prop50to64,
                                   factor = "Age 50 to 64"),
                        data.frame(POP.rec.cartogram,
                                   Proportion = Hamilton_CT$Prop65Plus,
                                   factor = "Age 65+"))
```

Create plot with frames:
```{r}
Cartograms.rec.plot <- ggplot(Cartograms.rec, aes(xmin = x - dx, xmax = x + dx, ymin = y - dy, ymax = y + dy, fill = cut_number(Proportion, 5))) + 
  geom_rect(aes(frame = factor)) +
  scale_fill_brewer(palette = "YlOrRd") +
  coord_fixed() +
  theme(legend.position = "bottom") +
  labs(fill = "Proportion")
```

Plot:
```{r}
plot <- Cartograms.rec.plot %>% animation_opts(frame = 1000, transition = 1000, easing = "elastic")
ggplotly(plot)
```